Grappa: Scalable Computing on Commodity Hardware \\for Irregular Applications

### 2012-10-16 doc/asplos13/rebuttal.txt

We thank our reviewers for their thoughtful comments.

Some reviewers (1, 4, 5) did not see a novel contribution in our
work. We believe that our system is the first runtime designed
specifically to tolerate latency by switching between thousands of
concurrent tasks per core. We are able to keep cores busy while
performing remote memory, synchronization, and work-stealing
operations. While the individual ideas used to build Grappa are
generally well-known, we know of no other runtime that combines these
ideas to trade latency for throughput.

Reviewer 3 takes issue with our characterization of our performance as
"on par" with that of the XMT. We are able to provide a simple,
general programming model similar to that of the XMT on
already-installed commodity clusters with at most a 2.5x slowdown,
even though the XMT has a 10x faster network injection rate and 20x
faster context switches than we do. We believe this slowdown is within
the realm of optimization---improvements to our network layer should
allow us to double the number of cores per node we use, and
improvements to our scheduler should allow us to increase the rate at
which each core injects messages into the network by ~30 percent.

Reviewer 2 asks about algorithmic modifications such as Beamer's BFS
transformation. Both the XMT and Grappa can benefit from such
transformations. Beamer's transformation increases BFS performance in
two ways: first, by reducing the number of edges traversed, and
second, by reducing the number of synchronizing operations
executed. Both the XMT and Grappa would benefit from the reduction in
edges traversed, but since Grappa's remote references are relatively
more costly than the XMT's, we would expect Grappa to benefit
more. Atomic operations on both platforms are no more expensive than
normal memory operations, so we would expect no significant benefit
from the reduction in synchronizing operations. Many transformations
also benefit by exploiting locality: Grappa provides mechanisms to do
so easily, while the hashed addressing of the XMT makes exploiting
locality difficult.

Reviewer 1 suggests comparing Grappa against other benchmark
implementations on commodity clusters in order to evaluate the benefit
of the programming model. In this work, we focused on providing a
similar programming model to the XMT, which is known to be simple
relative to distributed memory machines. We will explore programmer
productivity and compare against other commodity-cluster-based
implementations in future work.

Our evaluation was written in haste; in a final version we would
discuss key metrics including scheduling overhead and aggregator
performance limitations <Reviwer 2>.

We thank Reviewer 4 for expanding our understanding of the related
work. We will incorporate the suggestions.

[scalability?]

[citation for current memory hierarchy != irregular apps?]

### 2012-10-16 doc/asplos13/rebuttal.txt

We thank our reviewers for their thoughtful comments.

Some reviewers (1, 4, 5) did not see a novel contribution in our work. We
believe that our system is the first runtime designed specifically to tolerate
latency by switching between thousands of concurrent tasks per core. We are
able to keep cores busy while performing remote memory, synchronization, and
work-stealing operations. We aggregate small shared memory references into
larger network messages to better utilize the network. While the individual
ideas used to build Grappa are generally well-known, we know of no other
runtime that combines these ideas to trade latency for throughput.

Reviewer 3 takes issue with our characterization of our performance as "on
par" with that of the XMT. We are able to provide a simple, general
programming model similar to that of the XMT on already-installed commodity
clusters with at most a 2.5x slowdown, even though the XMT has a 10x faster
network injection rate and 20x faster context switches than we do. We believe
this slowdown is within the realm of optimization of the runtime features
already covered in the paper---improvements to our
network layer should allow us to double the number of cores per node we use,
and improvements to our scheduler should allow us to increase the rate at
which each core injects messages into the network by ~30 percent.  [We should
be precise about "similar programming model" to avoid more overstated claims,
because in Grappa's current form it is not quite the same. The main difference
is not a difficult one but it exists: in our benchmarks we take some locality into account: e.g. caching edgelist chunks]

@@

Reviewer 2 asks about algorithmic modifications such as Beamer's BFS
transformation. Both the XMT and Grappa can benefit from such
transformations. Beamer's transformation increases BFS performance in
two ways: first, by reducing the number of edges traversed, and
second, by reducing the number of synchronizing operations
executed. Both the XMT and Grappa would benefit from the reduction in
edges traversed, but since Grappa's remote references are relatively
more costly than the XMT's, we would expect Grappa to benefit
more. Atomic operations on both platforms are no more expensive than
normal memory operations, so we would expect no significant benefit
from the reduction in synchronizing operations. Many transformations
also benefit by exploiting locality: Grappa provides mechanisms to do
so easily, while the hashed addressing of the XMT makes exploiting
locality difficult.

Reviewer 1 suggests comparing Grappa against other benchmark
implementations on commodity clusters in order to evaluate the benefit
of the programming model. In this work, we focused on providing a
similar programming model to the XMT, which is known to be simple
relative to distributed memory machines. We will explore programmer
productivity and compare against other commodity-cluster-based
implementations in future work.

Our evaluation was written in haste; in a final version we would discuss key
metrics including scheduling overhead and aggregator performance limitations
<Reviwer 2>. The evaluation of the scheduler would break down scheduling
overhead into checking the polling thread, checking other
queues, doing load balancing, and context switch time, and compare the average overhead to the time a task executes between yields.

### 2012-10-16 doc/asplos13/rebuttal.txt

Reviewer 2 asks about algorithmic modifications such as Beamer's BFS
transformation. Both the XMT and Grappa can benefit from such
transformations. Beamer's transformation increases BFS performance in
two ways: first, by reducing the number of edges traversed, and
second, by reducing the number of synchronizing operations
executed. Both the XMT and Grappa would benefit from the reduction in
edges traversed, but since Grappa's remote references are relatively
more costly than the XMT's, we would expect Grappa to benefit
more. Atomic operations on both platforms are no more expensive than
normal memory operations, so we would expect no significant benefit
from the reduction in synchronizing operations. Many transformations
also benefit by exploiting locality: Grappa provides mechanisms to do
so easily, while the hashed addressing of the XMT makes exploiting
locality difficult.

Reviewer 1 suggests comparing Grappa against other benchmark
implementations on commodity clusters in order to evaluate the benefit
of the programming model. In this work, we focused on providing a
similar programming model to the XMT, which is known to be simple
relative to distributed memory machines. We will explore programmer
productivity and compare against other commodity-cluster-based
implementations in future work.

Our evaluation was written in haste; in a final version we would discuss key
metrics including scheduling overhead and aggregator performance limitations
<Reviwer 2>. The evaluation of the scheduler would break down scheduling
overhead into checking the polling thread, checking other
queues, doing load balancing, and context switch time, and compare the average overhead to the time a task executes between yields.
[TODO: what we can say in rebuttal without more data:
1. Why BC and BFS are worse while UTS is better than XMT. I think the story is
that our injection rate is not as good and UTS wins because of XMT not scaling
2. Why scaling of apps is BC < BFS < UTS.]
[TODO: Mention that we already have the means to give a more explanatory
evaluation: Grappa is instrumented and we have a lightweight tracing
infrastructure.]

### 2012-10-16 doc/asplos13/rebuttal.txt

We thank the reviewers for their thoughtful comments.

>>> What is the novelty of the work? (Reviewers 1, 4, 5)

We believe that our system is the first runtime designed specifically to
tolerate latency by switching between thousands of concurrent tasks per core.
We are able to keep cores busy while performing remote memory,
synchronization, and work-stealing operations. We aggregate messages from
unrelated threads into larger network messages to better utilize the network.
While the individual ideas used to build Grappa are generally well-known, we
know of no other runtime that combines these ideas to trade latency for
throughput to build a general-purposed parallel runtime for large-scale applications.

>>> Why a performance comparison with XMT? (Reviewer 3)

We are able to provide a simple, general programming model similar to that of
the XMT on ubiquitous commodity clusters with at most a 2.5x slowdown (and at
best TODO faster), even though the XMT has a 10x faster network injection rate
and 20x faster context switches than we do. We believe this slowdown is within
the realm of optimization of the runtime features already covered in the
paper---improvements to our network layer should allow us to double the number
of cores per node we use, and improvements to our scheduler should allow us to
increase the rate at which each core injects messages into the network by ~30
percent.

[This would just complicate things. I would leave it out... We should be
precise about "similar programming model" to avoid more overstated claims,
because in Grappa's current form it is not quite the same. The main difference
is not a difficult one but it exists: in our benchmarks we take some locality
into account: e.g. caching edgelist chunks]

>>> Why not compare with other benchmarks that running commodity clusters?
(Reviewer 1)

In this work, we focused on providing a similar programming model to the XMT,
which is known to be simple relative to distributed memory machines. We will
explore programmer productivity and compare against other
commodity-cluster-based implementations in future work.

[Can we say something more about how we would compare based on published results?]

>>> Evaluation section is unpolished and incomplete. (Reviewer 2)

We will cleanup and expand our evaluation. In the final version we will
discuss key metrics including scheduling overhead and aggregator performance
limitations. The evaluation of the scheduler would break down scheduling
overhead into checking the polling thread, checking other queues, doing load
balancing, and context switch time, and compare the average overhead to the
time a task executes between yields. We will expand the evaluation with a detailed study of where time goes, since Grappa has a lightweight tracing infrastructure --- we did not add this to the evaluation due to lack of time.

[You can explain all these points in the rebuttal. You just can't add more benchamrks or major results. TODO: what we can say in rebuttal without more data: 1. Why BC and BFS are
worse while UTS is better than XMT. I think the story is that our injection
rate is not as good and UTS wins because of XMT not scaling 2. Why scaling of
apps is BC < BFS < UTS.]

>>> What would be the impact of better algorithms (e.g., Beamer's BFS) in the
performance comparisons? (Reviewer 2)

Both the XMT and Grappa can benefit from such transformations. Beamer's
transformation increases BFS performance in two ways: first, by reducing the
number of edges traversed, and second, by reducing the number of synchronizing
operations executed. Both the XMT and Grappa would benefit from the reduction
in edges traversed, but since Grappa's remote references are relatively more
costly than the XMT's, we would expect Grappa to benefit more. Atomic
operations on both platforms are no more expensive than normal memory
operations, so we would expect no significant benefit from the reduction in
synchronizing operations. Many transformations also benefit by exploiting
locality: Grappa provides mechanisms to do so easily, while the hashed
addressing of the XMT makes exploiting locality difficult.

>>> Missing related work. (Reviewer 4)

Thanks for pointing to extra related work, we will add these to the final version.

### 2012-10-18 doc/asplos13/rebuttal.txt

Submitted:

We thank the reviewers for their thoughtful comments.

>>> What is the novelty of the work? (Reviewers 1, 3, 4, 5)

Conventional wisdom holds that to obtain performance on irregular applications one should minimize latency of individual memory and scheduling operations. In contrast, our approach delays and bundles these operations, exploiting the wealth of concurrency in the problem domain to overlap this additional latency with other computation and consequently achieve high overall throughput. We use this idea to obtain high throughput not only on remote memory operations, but also on remote synchronization and distributed work-stealing operations. We believe that our system is the first runtime designed specifically to tolerate multi-millisecond latencies by switching between millions of threads on hundreds to thousands of cores.

While the individual ideas used to build Grappa are generally well-known, we know of no other runtime that combines these ideas with this goal. We combine ideas from: (1) distributed shared memory, but we use it for applications with little locality, i.e., caches are not necessarily valuable; (2) network message aggregation, but we use it for unrelated remote memory accesses instead of explicit messages, as in prior work; and (3) distributed work stealing, but with asynchronous steals that are overlapped with other computation via multithreading.

The original goal of the project was to mimic the Cray XMT with commodity processors using a combination of hardware and software support. We thought hardware support would be necessary due to the vast difference in hardware capability. Each Cray XMT processor supports 1024 outstanding memory operations, while the processors in our cluster support a few dozen. The XMT processors have dedicated storage for thread contexts, whereas ours must share the data cache. The XMT includes extensive support for fine-grained synchronization, including full-bits and the ability to atomically increment a word anywhere in the system with a single instruction, while our cluster’s atomic operations are limited to a single node’s shared memory domain and may stall the pipeline to enforce the memory consistency model. We find it encouraging that Grappa performs as well as it does compared to the XMT without depending on specialized hardware.

>>> Why a performance comparison with XMT? (Reviewers 1, 3, 4)

In this work, we focused on providing a programming model similar to that of the XMT, which is known to be simple relative to that of distributed memory machines. We are able to provide this programming model on ubiquitous commodity clusters with at worst a 2.5x slowdown (and at best 3.2x speedup) on 16 nodes, even though the XMT has a 10x faster network injection rate and zero-cost context switches. By comparing against the XMT, we can hold the programming model constant and evaluate the performance difference between custom and commodity hardware under that model. Future work will compare Grappa against other commodity solutions in terms of performance as well as ease of programming.

>>> Is a 2.5x performance gap good or bad? What optimization opportunities exist? (Reviewers 1, 3)

We believe our slowdown with respect to the XMT is within the realm of optimization of the runtime features discussed in the paper. We were limited to less than half of the available cores in our system by contention in our network layer; improving that should allow us to double our network injection rate per node. Experiments with our context switch code suggest we can obtain a further ~30 percent increase in network injection rate per core.

Automatically tuning the runtime parameters is an interesting area of future work. We believe we can use core idleness and aggregated message sizes as signals to tune the worker count and aggregator delay, but it is not straightforward---low-work phases of execution as well as synchronization events may require temporary changes to these parameters to ensure rapid progress. A similar flavor of parameter exploration has been studied in previous work in task parallelism: e.g., parallel for-loops use granularity to trade off overhead of spawns, and adaptive work stealing schedulers switch between “work-first” and “help-first” policies.

>>> Evaluation section is unpolished and incomplete. What about efficiency, scheduling overhead, and aggregator performance limitations? (Reviewers 1, 2, 3, 4)

We will clean up and expand our evaluation. Grappa includes support for lightweight tracing and statistics collection, which was invaluable for debugging the performance of the runtime and benchmarks. We will use this to provide a detailed study of where time goes. An expanded evaluation will include analysis of scheduler overheads and execution efficiency, global memory and synchronization performance, and scaling challenges. We did not include this previously due to lack of time.

>>> What would be the impact of better algorithms (e.g., Beamer's BFS) in the performance comparisons? (Reviewer 2)

Since both the XMT and Grappa support a general programming model, both can benefit from such algorithmic modifications. Beamer's transformation increases BFS performance in two ways: first, by reducing the number of edges traversed, and second, by reducing the number of synchronizing operations executed. Both the XMT and Grappa would benefit from the reduction in edges traversed, but since Grappa's remote references are relatively more costly than the XMT's, we would expect Grappa to benefit more. Atomic operations on both platforms are no more expensive than normal memory operations, so we would expect no significant benefit from the reduction in synchronizing operations. Many transformations also benefit by exploiting locality: Grappa provides mechanisms to do so easily, while the hashed addressing of the XMT makes exploiting locality difficult.

We will consider implementing Beamer’s approach for the final version.

>>> Synchronization support (Reviewer 1)

Grappa does have a synchronization construct like Cilk_sync which blocks a task until all its subtasks have completed. This is used to implement Grappa's parallel loop support. Grappa also supports fine-grained synchronization primitives, such as condition variables, via delegate operations, without memory fences or expensive retries as in the XMT.

>>> Missing related work; reference for memory hierarchy challenges (Reviewers 4, 5)

Thanks for pointing to extra related work; we will add these to the final version.

We will also cite “Improving Memory Hierarchy Performance of Irregular Applications Using Data and Computation Reorderings” by Mellor-Crummey, Whalley, and Kennedy from IJPP 2001, which discusses the challenges of mapping irregular computation to commodity memory hierarchies.

Unsubmitted additional material for future use:

Finally, whether or not a purely software-based solution for commodity systems can offer customers of the Cray XMT a cost-effective alternative has been a long-standing question for those customers.  The answer has been anything but obvious:  that is why they have been eager to support this work.  The lack of immediacy in verdict is attributed to four technical challenges.

First was the question of whether or not a commodity processor could tolerate the latency of large systems without additional hardware support.  The obvious answer is “No:  each core is capable of issuing at most 10 memory operations before it stalls;  furthermore, each Nehalem memory controller can execute at most 32 memory operations (including software prefetches) before it stalls;  a Cray XMT processor with a network latency of only 2us and running at only 500MHz can issue up to 1024 memory references and still sometimes stalls;  clearly a 3GHz processor will need more than 32 outstanding memory references to tolerate the much greater latency of commodity networks.”  For quite some time, this response stood as the rationale for requesting that Intel processors be redesigned to get more references in flight.  Indeed, there have been proposals to use FPGAs as surrogate memory controllers as a means of increasing memory concurrency to remote memory locations.  Apparently -- surprisingly to some -- Grappa demonstrates that it is possible to effect the needed concurrency (several thousand remote memory references in flight) with sufficiently low overhead that a commodity system might in fact serve those customers’ purposes.

Second was the question of whether or not those thousands of threads’ contexts would “blow out cache”, defeating the purpose of latency tolerance.  Considering that a thousand threads might each execute many instructions each before any given thread that has yielded regains control of a core, one can imagine that both local and remote state that it had depended upon finding in cache to sustain performance might by then have been evicted.  Again, it is surprising that this is not the case:  that while there is some performance degradation due to such evictions, degradation does not sufficiently degrade overall performance to negate the gains of latency tolerance.

Third was the question of synchronization.  The Cray XMT architecture includes several instructions each of which perform some form of full-bit synchronization and an int_fetch_add operation.  With just one instruction, the XMT can increment a word of memory anywhere in the system.  Hardware supports retry operations when full-bit operations find a word in the “wrong” state.  Again, it is not obvious how to compete against this with a software-only system relying on processors whose atomic operations not only are limited in functionality but also induce fences, stalling the cpu pipelines until all outstanding stores are flushed.  And, again, some folks have proposed FPGA solutions.  Indeed, synchronization operations are dense in high-concurrency programs. We realized we could not use the x86 built-ins even for node-local data with any frequency without interfering with the very pipeline throughput we were aiming to improve through the use of concurrency.  But, somewhat surprisingly considering all the additional operations involved, our software-only solution in which every synchronizing and every global reference passes through a Grappa delegate appears sufficiently efficient to compete.

Fourth was the question of global memory model.  The Cray XMT guarantees sequential consistency for race free programs.  There are no caches so there are no coherency concerns.  The compiler injects a “lookahead” number into every memory operation that expresses how many more operations from that thread the pipeline may execute before it must wait for the memory operation to complete.  x86 processors have caches.  There is no hardware support for lookahead numbers.  We interweave threads outside the view of the compiler.  What the x86 does provide is total store order.  Putting all this together, it is not immediately clear what we get in the way of a consistency model.  However, what we believe we have is the same sequential consistency for race free programs as is offered by the Cray XMT, using off the shelf compilers and no hardware support specific to multiprocessor systems.  We believe without proof, yet, that we get this as a result of the total store order of individual processors combined with our delegation scheme that serializes all synchronizing and global requests.

In summary, while there may well be no novelty in any one technique, we believe there is novelty in the fact that Grappa works as well as it does, likely offering a satisfactory commodity-based alternative to Cray XMT customers, even though we still have a great deal of work ahead.

### 2012-10-18 doc/asplos13/rebuttal.txt

Fourth was the question of global memory model.  The Cray XMT guarantees sequential consistency for race free programs.  There are no caches so there are no coherency concerns.  The compiler injects a “lookahead” number into every memory operation that expresses how many more operations from that thread the pipeline may execute before it must wait for the memory operation to complete.  x86 processors have caches.  There is no hardware support for lookahead numbers.  We interweave threads outside the view of the compiler.  What the x86 does provide is total store order.  Putting all this together, it is not immediately clear what we get in the way of a consistency model.  However, what we believe we have is the same sequential consistency for race free programs as is offered by the Cray XMT, using off the shelf compilers and no hardware support specific to multiprocessor systems.  We believe without proof, yet, that we get this as a result of the total store order of individual processors combined with our delegation scheme that serializes all synchronizing and global requests.

In summary, while there may well be no novelty in any one technique, we believe there is novelty in the fact that Grappa works as well as it does, likely offering a satisfactory commodity-based alternative to Cray XMT customers, even though we still have a great deal of work ahead.

--Programming model 
As an example, we discuss what a UPC implementation of our
simplest benchmark, UTS-mem, (which only performs remote reads) would look
like. UTS-mem is a parallel search of an unbalanced tree in memory, assuming
no prior knowledge of the tree structure. UTS implementations use a work
stack, containing addresses of nodes to visit. The UPC Thread would deqeue the
global address and perform a high-latency remote read to get the node. With
the node information, it then performs another remote read to get the node's
edge list. The elements in the edge list can be added to the work stack. This
simple solution will spend most of the cpu's time waiting for remote reads. An
aggressive and fair implementation would need to overlap multiple reads for
vertices and edge lists.** This would involve a scheduling loop that manages
three structures: the work stack of unstarted vertices, the pool of
outstanding reads, and the pool of received vertices and edge lists.  The
Thread initiates nonblocking gets, polls for received messages (try_sync), and
executes received vertices and edge lists.  The need of the extra data
structures indicates that this level of communication overlap is too complex
to be extracted from the simple version by UPC's compiler optimizations. The
solution is further complicated by very large edge lists, which need to broken
down into constant size chunks to prevent serialization.  These optimizations
diverge from the UPC shared memory programming model.

There is also a
performance concern: the UPC solution is that just using nonblocking reads is
not sufficient. Commodity networks do not get good bandwidth for small
messages, and there is a limit to the number of outstanding RDMA operations.
The UPC solution could address this by aggregating reads to the same node
using active messages, but then you are building a general aggregation
mechanism that Grappa provides in the network layer.

**It might actually be possible to try manually unrolling the main loop of dequeue,read vertex,read edgelist,enqueue edges. By putting all the reads together in the unroll, the UPC compiler might be able to convert this gather into nonblocking ops. If so then it could be efficient in terms of the individual dependence from vertex to its edges. But all the work in one iteration may have to finish before the next iteration, in which case performance degradation is determined by the variability of non-uniform access times in the gather ops (dataflow within iteration, but synchronized iterations). Aggregation could be added by increasing the amount of unrolling in the loop.

### 2012-11-06 doc/asplos13/communication.tex

\section{Communication}

@@

In our experiments the vast majority of requests were smaller than
44 bytes, far too small to make
efficient use of the network. To make the best use of the network, we
must convert our small messages into large ones
(Section~\ref{sec:results-aggregation}).  When a task sends a message,
it is not immediately sent, but rather placed in a queue specific to the
destination.

There are three situations in which a queue of aggregated messages is
sent. First, each queue has a message size threshold, chosen to give
reasonable network performance 4096. If the size in
bytes of a queue is above the threshold, the contents of the queue are
sent immediately. Second, each queue has a wait time threshold
($\approx${1ms}). If the oldest message in a queue has been waiting
longer than this threshold, the contents of the queue are sent
immediately, even if the queue size is lower than the message size
threshold.  Third, queues may be explicitly flushed in situations where
the programmer wants to minimize the latency of a message at the cost of
bandwidth utilization.

>>>>

\section{Communication}\label{sec:communication}

@@

In our experiments the vast majority of requests were smaller than
44 bytes, far too small to make
efficient use of the network. To make the best use of the network, we
must convert our small messages into large ones.
When a task sends a message,
it is not immediately sent, but rather placed in a queue specific to the
destination.

There are three situations in which a queue of aggregated messages is
sent. First, each queue has a message size threshold of 4096 bytes, chosen to give reasonable network performance. If the size in
bytes of a queue is above the threshold, the contents of the queue are
sent immediately. Second, each queue has a wait time threshold
($\approx${1ms}). If the oldest message in a queue has been waiting
longer than this threshold, the contents of the queue are sent
immediately, even if the queue size is lower than the message size
threshold.  Third, queues may be explicitly flushed in situations where
the programmer wants to minimize the latency of a message at the cost of
bandwidth utilization.
